Data exploration through dot-driven development\footnote{This work was supported by Google Digital News Initiative.}

### 2016-12-31 pivot/paper.tex
Tracking columns, part one

The core language uses type providers, but we employ them in a novel way -- rather than providing 
types with members for accessing data, we provide types with members that allow the user to access 
data, but also compose complex queries using only member access (``dot''). This lets us recreate 
functionality that usually requires complex type systems (row polymporphism, phantom types, 
dependent typing) in an extremely simple object-based language.

@@

\section{Introduction -- Simplifying data exploration}
\label{sec:intro}

@@

\section{Background -- Type providers}
\label{sec:tps}

@@

\section{Analysis -- Simplifying data scripting}
\label{sec:analysis}

@@

proponents of static typing would correctly
point out that both member names and column names can be tracked by a more sophisticated type
system. However, powerful types usually increase the complexity of the programming language.

@@

\subsection{Using row types and phantom types}
\label{sec:columns-row}

Phantom types + row types

\begin{equation*}
  \tau = [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n] \lsep \tau_\ident{lbl}
\end{equation*}

\begin{equation*}
\inference
  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
  {\Gamma \vdash e.\qident{drop columns}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{sort}}}
\end{equation*}

\begin{equation*}
\inference
  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{sort}}}
  {\Gamma \vdash e.\qident{then}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
\end{equation*}

\begin{equation*}
\inference
  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{sort}}}
  {\Gamma \vdash e.\qident{drop~$f_i$}: [f_1\!:\!\tau_1, \ldots, f_{i-1}\!:\!\tau_{i-1}, f_{i+1}\!:\!\tau_{i+1}, \ldots, f_n\!:\!\tau_n ]_{\ident{sort}}}
\end{equation*}

\subsection{Using type providers}
\label{sec:columns-tp}

>>>>

The core language uses type providers, but we employ them in a novel way -- rather than providing 
types with members for accessing data, we provide types with members that allow the user to access 
data, but also compose complex queries using only member access (``dot''). This lets us recreate 
functionality that usually requires complex type systems (row polymporphism, type state, 
dependent typing) in an extremely simple object-based language.

@@

\section{Introduction}
\label{sec:intro}

@@

\section{Using type providers in a novel way}
\label{sec:tps}

@@

\section{Simplifying data scripting languages}
\label{sec:analysis}

@@

The last difficulty with data scripting discussed in Section~\ref{sec:analysis-why} is that pandas 
(and other data exploration libraries, even for statically-typed languages) track column names as 
strings at runtime, making code error-prone and auto-complete on column names difficult to provie.
Proponents of static typing would correctly point out that column names and their types can be
tracked by a more sophisticated type system.

In this section, we discuss our approach -- we track column names statically using a mechanism
that is inspired by row types and type state (Section~\ref{sec:columns-row}), however we 
embed the mechanism using type providers into a simple nominal type system (Section~\ref{sec:columns-tp}).
This way, the host language for the pivot type provider can be extremely simple -- and indeed, the
mechanism could be added to languages such as Java or TypeScript with minimal effort.

@@

\begin{figure}

\begin{equation*}
\inference[(\emph{drop-start})~]
  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
  {\Gamma \vdash e.\qident{drop columns}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
\end{equation*}

\begin{equation*}
\inference[(\emph{drop-col})~]
  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
  {\Gamma \vdash e.\qident{drop~$f_i$}: [f_1\!:\!\tau_1, \ldots, f_{i-1}\!:\!\tau_{i-1}, f_{i+1}\!:\!\tau_{i+1}, \ldots, f_n\!:\!\tau_n ]_{\ident{drop}}}
\end{equation*}

\begin{equation*}
\inference[(\emph{drop-then})~]
  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
  {\Gamma \vdash e.\qident{then}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
\end{equation*}

\caption{Tracking available column names with row types and type state}
\label{fig:fancy-types}
\end{figure}

\subsection{Using row types and type state}
\label{sec:columns-row}

There are several common data transformations that modify the structure of the data set and affect
what columns (and of what types) are available. When grouping and aggregating data, the resulting
data set has columns depending on the aggregates calculated. Another simpler operation is adding
or removing a column from the data set. For example, given the Olympic medals data set, we can 
drop games and year as follows:
%
\begin{equation*}
  \begin{array}{l}
    \ident{olympics}.\qident{drop columns}.\qident{drop Games}.\qident{drop Year}.\ident{then}
  \end{array}
\end{equation*}

\noindent
Operations on the type of rows in the data set can be captured using row types [X]. In addition, 
we need to annotate type with a form of type state [Y] to restrict what operations are available. 
When dropping columns, we first access the \qident{drop columns} member, which sets the state to
a state where we can drop individual columns using \qident{drop $f$}. The \ident{then} member can 
then be used to complete the operation and choose another transformation.

To illustrate tracking of columns using row types and type state, consider a simple language with
variables (representing external data sources) and member access. Types can be either primitive 
types $\alpha$, types annotated with a type state \ident{lbl} or row type with fields $f$:
%
\begin{equation*}
\begin{array}{rcl}  
e&=&v \lsep e.N\\
\tau&=&\alpha \lsep \tau_\ident{lbl} \lsep [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n] 
\end{array}  
\end{equation*}

\noindent
Typing rules for members that are used to drop columns are shown in Figure~\ref{fig:fancy-types}.
When \qident{drop columns} is invoked on a record, the type is annotated with a state \ident{drop}
(\emph{drop-start}) indicating that individual columns may be dropped. The \ident{then} 
operation (\emph{drop-then}) removes the state label. Individual members can be removed using
\qident{drop $f_i$} and the (\emph{drop-col}) rule ensures the dropped column is available in the
input row type and removes it.

Other data transformations could be type checked in a similar way, but there are two drawbacks.
First, row types and type state (although relatively straightforward) make the host language more
complex. Second, rules such as (\emph{drop-col}) make auto-completion more difficult, because the
editor needs to understand the rules and calculate what members may be invoked. This is a distinct
operation from type checking and type infernece (which operate on complete programs) that needs 
to be formalized and implemented.

\subsection{Using the pivot type provider}
\label{sec:columns-tp}

If $\Gamma\vdash e:[f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]$ using row types and
$\Gamma\vdash e:C$ then $\ident{fields}(C) = \{ f_1\mapsto\tau_1 \}$
\begin{equation*}
\end{equation*}
